---
title: "Pella Tomlinson Exercise"
author: "Pamela J. Woods"
date: "2024-01-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction

Google Pella Tomlinson, describe what it is. <https://haddonm.github.io/URMQMF/surplus-production-models.html>

## Methods

Describe the exercise.

## Results and Discussion

R code and a plot.

Below, copy and paste the Schaefer model, then modify it to be Pella Tomlinson.

```{r model}

#hint: to avoid overwriting results later, save the vectors you want to plot with different names than used for the Schaeffer model
    #eff_p<-seq(0,2,0.1) 
    #YY_p[j] <- 
    #XX_p[j] <-   
  
```

Below, copy and paste the Schaeffer model. Make the vectors used for recording results are not the same as above.

```{r model2}
  #eff<-seq(0,2,0.1) # values of effort to loop over
  #YY<-vector(mode = "numeric", length = length(eff))
  #XX<-vector(mode = "numeric", length = length(eff))
  
  
```

Plot effort x Yield from Pella Tomlinson.

```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
