---
title: "Pella Tomlinson Exercise"
author: "Pamela J. Woods"
date: "2024-01-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction

Google Pella Tomlinson, describe what it is. <https://haddonm.github.io/URMQMF/surplus-production-models.html>

## Methods

Describe the exercise.

## Results and Discussion

R code and a plot.

Below, copy and paste the Schaefer model, then modify it to be Pella Tomlinson.

```{r model}
  eff_p<-seq(0,2,0.1) # values of effort to loop over
  YY_p<-vector(mode = "numeric", length = length(eff_p))
  XX_p<-vector(mode = "numeric", length = length(eff_p))
  
  Catchability<-0.1
  TT<-1:500
  KK<-2000
  PP0<-1; rr<- 0.2
  p <- 1.5
  
  
  for(j in 1:length(eff_p)){
    
    PP<-vector(mode = "numeric", length = 500); 
    DD<-vector(mode = "numeric", length = 500);
    CC<-vector(mode = "numeric", length = 500); 
    
    Effort<- eff_p[j] 
    
    DD0<-rr/p*PP0*(1-(PP0/KK)^p)
    CC0<-PP0*Effort*Catchability; 
    PP[1]<-PP0 + DD0 - CC0
    
    for(i in 2:500){ 
      
      DD[i-1]<- rr*PP[i-1]/p*(1-(PP[i-1]/KK)^p)
      CC[i-1]<- PP[i-1]*Effort*Catchability
      PP[i]<- PP[i-1] + DD[i-1] - CC[i-1]
      
    } 
    
    DD[500]<- rr*PP[500]/p*(1-(PP[500]/KK)^p)
    CC[500]<- PP[500]*Effort*Catchability
    
    #SAVE RESULTS: LAST (EQUILIBRIUM) VALUE OF CATCH AND LAST VALUE OF POPULATION 
    YY_p[j] <- CC[500] # Equilibrium Yield
    XX_p[j] <-  PP[500] #or PP[500] # Equilibrium Population B
    
  }
  
  
  p1 <- tibble(XX_p,YY_p,eff_p) %>% ggplot(aes(eff_p, YY_p)) + geom_point() + ylab("Yield") + xlab( "Effort")
  p2 <- tibble(XX_p,YY_p,eff_p) %>% ggplot(aes(eff_p, XX_p)) + geom_point() + ylab("Population B") + xlab( "Effort")
  p3 <- tibble(XX_p,YY_p,eff_p) %>% mutate(CPUE = YY_p/eff_p) %>% ggplot(aes(eff_p, CPUE)) + geom_point() + xlab("Effort") 
  
  gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
  
  #If cost linearly increases with effort...
  Cost<-eff_p*40
  #Q - then what is maximum economic yield (MEY)? [Hint: MEY is the yield with the greatest profit; that is, the greatest difference between Yield and cost]
  tibble(YY_p, eff_p, Cost) %>% ggplot(aes(eff_p, YY_p)) + geom_point() + ylab("Yield") + xlab("Effort") + geom_point(aes(eff_p, Cost), shape = 2)

  # MSY = K*r/4
  # Nmsy = K/2
  
  eff_p[which.max(YY_p)] #MSY
  
  eff_p[which.max(YY_p - Cost)] #MEY
```

Below, copy and paste the Schaeffer model. Rename the vectors for Saving Results.

```{r model2}
  eff<-seq(0,2,0.1) # values of effort to loop over
  YY<-vector(mode = "numeric", length = length(eff))
  XX<-vector(mode = "numeric", length = length(eff))
  
  Catchability<-0.1
  TT<-1:500
  KK<-2000
  PP0<-1; rr<- 0.2
  
  for(j in 1:length(eff)){
    
    PP<-vector(mode = "numeric", length = 500); 
    DD<-vector(mode = "numeric", length = 500);
    CC<-vector(mode = "numeric", length = 500); 
    
    Effort<- eff[j] 
    
    DD0<-rr*PP0*(1-PP0/KK)
    CC0<-PP0*Effort*Catchability; 
    PP[1]<-PP0 + DD0 - CC0
    
    for(i in 2:500){ 
      
      DD[i-1]<- rr*PP[i-1]*(1-PP[i-1]/KK)
      CC[i-1]<- PP[i-1]*Effort*Catchability
      PP[i]<- PP[i-1] + DD[i-1] - CC[i-1]
      
    } 
    
    DD[500]<- rr*PP[500]*(1-PP[500]/KK)
    CC[500]<- PP[500]*Effort*Catchability
    
    #SAVE RESULTS: LAST (EQUILIBRIUM) VALUE OF CATCH AND LAST VALUE OF POPULATION 
    YY[j] <- CC[500] # Equilibrium Yield
    XX[j] <-  PP[500] #or PP[500] # Equilibrium Population B
    
  }
  
  
  p1 <- tibble(XX,YY,eff) %>% ggplot(aes(eff, YY)) + geom_point() + ylab("Yield") + xlab( "Effort")
  p2 <- tibble(XX,YY,eff) %>% ggplot(aes(eff, XX)) + geom_point() + ylab("Population B") + xlab( "Effort")
  p3 <- tibble(XX,YY,eff) %>% mutate(CPUE = YY/eff) %>% ggplot(aes(eff, CPUE)) + geom_point() + xlab("Effort") 
  
  gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
  
  #If cost linearly increases with effort...
  Cost<-eff*40
  #Q - then what is maximum economic yield (MEY)? [Hint: MEY is the yield with the greatest profit; that is, the greatest difference between Yield and cost]
  tibble(YY, eff, Cost) %>% ggplot(aes(eff, YY)) + geom_point() + ylab("Yield") + xlab("Effort") + geom_point(aes(eff, Cost), shape = 2)

  # MSY = K*r/4
  # Nmsy = K/2
  
  eff[which.max(YY)] #MSY
  
  eff[which.max(YY - Cost)] #MEY
  
```

Plot effort x Yield from Pella Tomlinson.

```{r pressure, echo=FALSE}
p1 + geom_point(aes(eff_p, YY_p), shape=2, data = tibble(eff_p, YY_p), color = 'red' )
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
