---
title:  "Gró-FTP - Stock assessment line - Simple pop. models"
format: html
editor: visual
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(palmerpenguins)
```

# Simple population models

# Basic models

## Exponential models

Wikipedia example: if a species of bacteria doubles every 10 minutes, starting with 1 bacteria, how many bacteria would there be after 1 hour?

There are 3 pieces of information here, which we can represent with mathematical notation: $N_0$ = initial population = 1 R = rate of increase = 2 per 10 minutes (or 2/(10 mins/60 mins) in per-hour units) T = time step = 10 minutes (or 10 min/60 mins in per-hour units) t = total number of time steps = 60 mins/10 mins (or 1 hour/ (10/60 hour)) = 6

Here is an equation that describes the process of increasing over time:

$N_t = N_0*R^t$

$N_6 = 1*2^6 = 64$

After one hour, or 6 ten-minute intervals, there would be sixty-four bacteria.

RExcercies...$N_t = N_0*R^t = R*N_{t-1} = N_{t-1} + r*N_{t-1}$

Therefore, if $N_t = N_{t-1} + r*N_{t-1}$, then $dN/dt = r*N$

### Exercise 1

```{r}
#Below, TT and PP are storage vectors for your time step (TT) 
#and your population abundance measure at that time step. So TT[1] shows year 1 
#as your first time step; PP[1] shows your population abundance at TT[1]. 
PP<-vector(mode = "numeric", length = 5); RR<-exp(1); PP0<-1
TT<-1:5
PP[1]; TT[1]

#RR and PP0 above are the 2 parameters included in this model (in contrast to storage). 
#The 2 parameters affect the values resulting in PP, based on TT). 
#Therefore, they affect the shape of your population growth function. 
#TT is input, PP is output, and the PP output values depend on parameters RR and PP0.
# See how equations in slides are translated into the model below. Also try to keep 
# track of your parameters versus storage components as the models presented become 
# more complex.

#This is the model - it updates your expected population abundance at each of the 
#5 time step, assuming the population grows exponentially.
for(i in 1:5){ PP[i]<-PP0*exp(TT[i])}

#Note that the model can also be written as below with the same result.
for(i in 1:5){ PP[i]<-PP0*RR^TT[i]}
PP
par(mfrow = c(1,3))
plot(PP~TT, xlab = "Time", ylab = "Population N")

#### Still Other ways to write the exponential model ... 
PP<-vector(mode = "numeric", length = 5); PP0<-1
TT<-1:5
PP[1]<-PP0*RR
for(i in 2:5){ PP[i]<-RR*PP[i-1]} 
PP
plot(PP~TT, xlab = "Time", ylab = "Population N") # same graph

# If RR = 1 + rr, then this is the same as PP[i]<-(1 + rr)*PP[i-1] = PP[i-1] + rr*PP[i-1].
#  Change the code above to the last form to show this: 

PP<-vector(mode = "numeric", length = 5); PP0<-1
TT<-1:5
rr <- RR - 1
PP[1]<-PP0 + rr*PP0
PP[1]<-PP0*(1 + rr)
for(i in 2:5){ PP[i]<-PP[i-1] + rr*PP[i-1]} 
for(i in 2:5){ PP[i]<-PP[i-1]*(1 + rr)} 
PP
plot(PP~TT, xlab = "Time", ylab = "Population N") # same graph



```

## Logistic growth

$N_t = N_{t-1} + r*N_{t-1}*(1-N_{t-1}/K)$

-   What is the new parameter?

-   What is dN/dt?

-   What differs from exponential growth?

### Exercise

```{r}
# Add density dependence into the code below to reflect logistic growth. 

PP<-vector(mode = "numeric", length = 10); PP0<-2; rr<-1 ;
TT<-1:10
KK<-10
PP[1]<-PP0+rr*PP0*(1-PP0/KK)
for(i in 2:10){ PP[i]<-PP[i-1] + rr*PP[i-1]*(1-PP[i-1]/KK)} 
PP
plot(PP~TT, xlab = "Time", ylab = "Population N", ylim = c(0,12)) 

# Now Change rr and K values above to see how they affect the model 
#Q2.0: Try KK between values of 10 and 70. Why does it control?

PP<-vector(mode = "numeric", length = 10); PP0<-2; rr<-1 ;
TT<-1:10
KK<-70
PP[1]<-PP0+rr*PP0*(1-PP0/KK)
for(i in 2:10){ PP[i]<-PP[i-1] + rr*PP[i-1]*(1-PP[i-1]/KK)} 
PP
plot(PP~TT, xlab = "Time", ylab = "Population N") 


#Q2.1: Set KK at 50 and rr at 2. Why do you get strange shapes at higher rr values?

PP<-vector(mode = "numeric", length = 1500); PP0<-2; rr<-2 ;
TT<-1:1500
KK<-50
PP[1]<-PP0+rr*PP0*(1-PP0/KK)
for(i in 2:1500){ PP[i]<-PP[i-1] + rr*PP[i-1]*(1-PP[i-1]/KK)} 
PP
plot(PP~TT, xlab = "Time", ylab = "Population N") 


```

## Further considerations

-   What biological and ecological factors contribute to r and K?
-   What population characteristic is the most important for natural resource management?
-   How to detect this without biological/ecological data?
-   Draw a graph of how population growth rate ($dN/dt$) changes over time for the case studies of 1) exponential and 2) logistic growth. How does the density dependent term $(1-N/K)$ change over time?
-   What species would be difficult to represent with this?
