---
title: "Pella Tomlinson Exercise"
author: "Pamela J. Woods"
date: "2024-01-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction

Google Pella Tomlinson, describe what it is. This is a good place to start: <https://haddonm.github.io/URMQMF/surplus-production-models.html>

## Methods

Describe the exercise.

## Results and Discussion

R code and a plot.

Below, copy and paste the Schaefer model, then modify it to be Pella Tomlinson. Rename the vectors so that they will not be overwritten by code in the next R chunk.

```{r model}
    #p1_p <- tibble(XX,YY_p,eff_p) %>% ggplot(aes(eff_p, YY_p)) + geom_point() + ylab("Yield") + xlab( "Effort")
```

Below, copy and paste the Schaeffer model again, but you don't need to change anything.

```{r model2}
  
      #p1 <- tibble(XX,YY,eff) %>% ggplot(aes(eff, YY)) + geom_point() + ylab("Yield") + xlab( "Effort")
```

Plot effort x Yield from Pella Tomlinson.

```{r pressure, echo=FALSE}
#p1 + geom_point(aes(eff_p, YY_p), shape=2, data = tibble(eff_p, YY_p), color = 'red' )
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
